---
title: "CGPfunctions"
author: "Chuck Powell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette 
vignette: >
  %\VignetteIndexEntry{CGPfunctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


A package that includes functions that I find useful for teaching statistics as well as actually practicing the art.  They typically are not "new" methods but rather wrappers around either base R or other packages and concepts I'm trying to master.  Currently contains:

- `Plot2WayANOVA` which as the name implies conducts a 2 way ANOVA and plots the results using `ggplot2`
- `neweta` which is a helper function that appends the results of a Type II eta squared calculation onto a classic ANOVA table

## Vignette Info

I need more text here.

## Examples

`neweta(aov(mpg~am*cyl, mtcars))`

$$\eta^2 = \frac{SS_{factor(term)}}{SS_{total}}$$.

```{r, echo=FALSE, results='asis'}
knitr::kable(CGPfunctions::neweta(aov(mpg~am*cyl, mtcars)))
```

`Plot2WayANOVA(mpg~am*cyl, mtcars)`

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
xxx<-CGPfunctions::Plot2WayANOVA(mpg~am*cyl, mtcars)
```

## Credits

Many thanks to Dani Navarro and the book  > ([Learning Statistics with R](http://www.compcogscisydney.com/learning-statistics-with-r.html)) whose etaSquared function  was the genesis of `neweta`.


> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

