---
title: "Using Plot2WayAnova"
author: "Chuck Powell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Plot2WayAnova}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background
The CGPfunctions package includes functions that I find useful for teaching
statistics especially to novices (as well as an opportunity to sharpen my own R
skills). I only write functions when I have a real need -- no theory -- just
help for actually practicing the art. They typically are not "new" methods but
rather wrappers around either base R or other packages and are very task
focused. This vignette covers one function from the package that tries to help
users (especially students) do one thing well by pulling together pieces from a
variety of places in `R`. `Plot2WayANOVA`, which as the name implies conducts a
2 way ANOVA and plots the results. I always try and find the right balance
between keeping the number of dependencies to a minimum and not reinventing the
wheel and writing functions that others have done for me. The function makes use
of the following non base r packages.

 - `ggplot2` as the work horse for all the actual plotting 
 - `car` for it's ability to compute Type II sums of squares, we'll address why
that's important in more detail later in the scenario. We'll also make use of
it's `leveneTest`.
 - `sjstats` which takes out ANOVA table and gives us other important
information such as the effect sizes ($\eta^2$ and $\omega^2$ ) through use of
its `anova_stats` function. Prior to this version I had been using my own local
function but this runs rings around what I could do.
 - `broomExtra::glance` will also help us grab very important results like $R^2$
and display them

## Vignette Info

The ANOVA (Analysis of Variance) family of statistical techniques allow us to
compare mean differences of one outcome (dependent) variable across two or more
groups (levels) of one or more independent variables (factor). It is also true
that ANOVA is a special case of the GLM or regression models so as the number of
levels increase it might make more sense to try one of those approaches. The 2
Way ANOVA allows for comparisons of mean differences across 2 independent
variables `factors` with a varying numbers of `levels` in each `factor`.

If you prefer a more regression based approach with a very similar plotted
result I highly recommend the `interactions` package which I was unaware of
until just recently.  It is [available through CRAN](https://cran.r-project.org/web/packages/interactions/index.html).

The `Plot2WayANOVA` function conducts a classic analysis of variance (ANOVA) in
a sane and defensible manner, not necessarily the only one. It's real strength
(*I hope*) lies in the fact that it is pulled together in one function and more
importantly allows you to visualize the results concurrently with no additional
work.

## Scenario and data

Imagine that you are interested in understanding whether a car's fuel efficiency
(mpg) varies based upon the type of transmission (automatic or manual) and the
number of cylinders the engine has. Let's imagine that the `mtcars` data set is
actually a random sample of 32 cars from different manufacturers and use the
mean `mpg` grouped by `am` and `cyl` to help inform our thinking. While we
expect variation across our sample we're interested in whether the differences
between the means by grouping of transmission type and cylinders is
significantly different than what we would expect in random variation across the
data.

In simplistic terms we want to know whether `am` matters, `cyl` matters or if it
depends on the interaction of the two. It's this interaction term that typically
confuses novices or is difficult to "see". That's where a good interaction graph
can hopefully play a key role, and that's what the `Plot2WayANOVA` focuses on.

There's no lack or tools or capabilities in base R or in the many packages to do
this task. What this function tries to do is pull together the disparate pieces
with a set of sane defaults and a simple interface to work with it. At its
simplest you would require the library and then enter this command:

`Plot2WayANOVA(mpg ~ am * cyl, mtcars)` which lays our question out in R's
vernacular with a formula and a dataframe. Optionally we can specify a different
confidence level and choose a line or a bar graph.

"Under the hood", however there's a lot of nice features at work.

 1. Some basic error checking to ensure a valid formula and dataframe. The
    function accepts only a fully crossed formula to check for an interaction 
    term 
 1. It ensures the dependent (outcome) variable is numeric and that the two
    independent (predictor) variables already are or can be coerced to factors
    – the user is warned on the console if there are problems.
 1. A check is conducted to see if any of the variables of interest have 
    missing cases – the user is warned on the console if there are problems.
 1. Balance is checked, that is whether the number of observations per cell
    when we cross `am` and `cyl` is equal. There's some debate as to how much
    imbalance is permissible. But you'll be given fair warning if there is any.
 1. In addition to the classic ANOVA table information available in `aov` or
    `Anova` you'll be presented with information about effect sizes like eta 
    squared $\eta^2$. They're calculated and appended as additional columns. 
    If you're unfamiliar with them and want to know more especially where the
    numbers come from I recommend a good introductory stats text. I recommend
    *Learning Statistics with R*
    [LSR](http://dj-navarro.appspot.com/lsr/lsr-0.5.1.pdf) see Table 14-1 
    on page 432.
 1. A summarized table of means, standard deviations, standard errors of the
    means, confidence intervals, and group sizes for each of the crossed
    combinations in our example that's 6 groupings 3 levels of cylinder and 2 levels
    of automatic or manual.
 1. Some measures of overall fit including $R^2$
 1. The Homogeneity of Variance assumption is tested with Brown-Forsythe 
 1. The normality assumption is tested with Shapiro-Wilk

## Installation

```{r eval = FALSE}
# Install from CRAN
install.packages("CGPfunctions")

# Or the development version from GitHub
# install.packages("devtools")
devtools::install_github("ibecav/CGPfunctions")
```

then load the library.

```{r LoadLibrary, warning = FALSE}
library(CGPfunctions)
```

## Example of using the function

The only two required parameters are a formula and a dataframe (like) object. If
we run the function in its simplest form here's what we get.

```{r Plot2WayANOVA, echo=TRUE, message=TRUE, warning=FALSE, fig.width=6.0, fig.height=3.5}
Plot2WayANOVA(formula = mpg ~ am * cyl, dataframe = mtcars)
```

In the console you'll receive a bunch of messages chronicling your progress
and any diagnostic information. In this case `am` and `cyl` are being coerced to
factors and you're being prompted to make sure that's what is intended. 

Next you receive a warning because you have a very unbalanced design. There are
only two 8 cylinder cars with a manual transmission and 12 with automatics.
Whereas there are eight 4 cylinders with manual and only three that are
automatics. Imbalance in our design worries us for two reasons. One is that it
causes a lack of statistical power and creates some math challenges in deciding
how to divide up the sums of the squared differences. This dataset causes the
more troublesome worry. Are the number of cylinders and manual versus automatic
related systematically which would call are whole design into question.  Make sure
you can answer questions about which is at work here, or make sure you have a 
balanced design.

A table follows that is intended to summarize the findings.  We'll discuss it 
more later when we examine the plot.

The overall measures can be very handy for comparing numerous models. For example
how does the AIC number change if we were to eliminate `am` ?

The table of group means is useful for looking at summary by group.  Want the best 
gas mileage buy a 4 cylinder manual car.


## Credits

Many thanks to Danielle Navarro and the book > ([Learning Statistics with
R](http://www.compcogscisydney.com/learning-statistics-with-r.html)).

